Going Beyond Ideal Point Points: Modeling Measurement Model Measurement Error

### 2023-09-15 paper.qmd
fixing some typos
penultimate draft!!

Theory-testing research which uses estimates from measurement models typically reduces the posterior distributions down to a single value. In the case of continuous variables, researchers select some statistic of central tendency from each posterior distribution to use in subsequent analyses, such as the mean, median, or mode. This practice necessarily discards information from the full distribution. @fig-measurement-model show four hypothetical posterior distributions that may arise from a Bayesian measurement model. Despite all having the same mean of zero, higher order moments such as variance (top-right), skew (bottom-left), and kurtosis (bottom-right) can create distributions which vary to a large extent.

@@

N <- 100000
tibble(x = c(rnorm(N, sd = 1),
             rnorm(N, sd = 2),
             sn::rsn(N, alpha = -4, omega = 2, xi = 1.5),
             sn::rst(N, xi = -2.75, omega = 2, alpha = 4, nu = 1)),
       y = c(rep("A\nNormal(0, 1)", N), 
             rep("B\nNormal(0, 2)", N), 
             rep("C\nSkew Normal(1.5, 2, -4)", N), 
             rep("D\nSkew t(-2.75, 2, 4, 1)", N))) |> 
  ggplot(aes(x = x)) +
  geom_histogram(color = "white", fill = "darkcyan", alpha = .5, bins = 50) +
  geom_vline(xintercept = 0, linetype = "twodash", color = "black", linewidth = .7) +
  # geom_text(aes(label = y), x = -4, y = 0) +
  facet_wrap(~ y, nrow = 2) +
  xlim(-5, 5) +
  theme_ggdist() +
  labs(y = "", x = "", caption = "Four different measurement model posterior distributions with mean zero") +
  theme(axis.line.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        text = element_text(family = "serif"))
```

@@

How can researchers avoid the issues highlighted above? In short, the measurement process and theory-testing procedure should happen simultaneously in a single model. This is handled straightforwardly using the Bayesian statistical framework, which can easily treat parameters and data interchangeably [@mcelreath2020]. We start by specifying the full measurement model, whose posterior distributions for each observation's value of the latent variable are then used as data in the theory-testing model. The stylized version of this joint model is shown in @eq-ex-equation, where $g(\cdot)$ is the measurement model which produces posterior estimates of the latent variable, $\theta_i$ for each observation based on some training data $y^*$. The posterior estimates for $\theta_i$ from the measurement model $g(\cdot)$ are then used as data in the theory-testing model $f(\theta_i)$ using the outcome of interest $y$.

$$
\begin{aligned}
  y_i &\sim f(\theta_i) \\
  y^*_i &\sim g(\theta_i)
\end{aligned}
$$ {#eq-ex-equation}

@@

The method developed in this paper overcomes the two problems outlined above by simplifying the measurement model step, $y^*_i \sim g(\cdot)$ in the joint model. Rather than estimating the latent variable from scratch, I take the posterior distributions already provided from previously fitted measurement models and use those as approximations in the full joint model. The measurement model $g(\cdot)$ becomes a probability distribution function with distributional parameters according to maximum likelihood estimates of the posterior. So if the posterior distribution of the measurement model appears normal, we would use $\bar{\theta}_i \sim N(\theta_i, \sigma^2_{\theta[i]})$. The values $\bar{\theta}_i$ and $\sigma^2_{\theta[i]}$ are estimated from the measurement model's posterior distribution, which allows the true, unobserved, value of the latent variable $\theta_i$ to be estimated for each observation. If the posterior distributions from the measurement model appear skewed, or have thicker tails than a normal distribution, the distributional parameters for these distributions can be used instead. These simplifications faithfully propagate the uncertainty in the outputs of the measurement model to the theory-testing model, while also being computationally tractable and straightforward to implement.

@@

One of the most common measurement models in political science is the Bayesian Item-Response Theory (IRT) model used to measure the ideological leanings of political actors [@clinton2004; @bafumi2005]. These models assume that political ideology is a latent characteristic that lies on a single left-right dimension. Observed actions, such as voting on legislation, are used as training data ($y^*$ in @eq-ex-equation) to produce a posterior distribution of continuous values for each actor (e.g. member of Congress) on this left-right scale.

Let's say we want to estimate the effect of legislator ideology, $\theta$, on some outcome, $y$. Using the format I developed in @eq-ex-equation, @eq-irt shows an example joint measurement and theory-testing model to answer this question using ideology estimates from an IRT model. The measurement model (bottom) predicts whether a legislator voted yes or no on a piece of legislation, $y^*$, using the traditional 2-parameter IRT equation $\Phi(\gamma_j\theta_i + \xi_j)$. Estimates of the parameters $\theta_i$ from this model, are then used as data in the linear regression theory-testing model (top) to estimate $\beta_1$---the coefficient of substantive interest.

$$
\begin{aligned}
  y_i &\sim \text{Normal}(\beta_0 + \theta_i \beta_1, \sigma^2) \\
  y^*_{ij} &\sim \text{Bernoulli}[\Phi(\gamma_j\theta_i + \xi_i)] \\
\end{aligned}
$$ {#eq-irt}

@@

$$
\begin{aligned}
  y_i &\sim \text{Normal}(\beta_0 + \beta_1 \bar{\theta}_i, \sigma^2) \\
  \bar{\theta}_i &\sim \text{Normal}(\theta_i, \sigma^2_{\theta [i]})
\end{aligned}
$$ {#eq-irt-me}

@@

Using the logic of causal graphs, I will discuss two types of measurement model measurement error, and how the joint measurement theory-testing procedure laid out in @eq-ex-equation helps fix them. First I will consider random, or classical, measurement error. In this case the joint model will (in most cases) provide researchers with extra statistical power to test their theory. Then we will look at non-random measurement error scenarios, in which the measurement model error introduces confounding in the theory-testing model. I will show how the joint model from @eq-ex-equation ameliorates this confounding bias. For both types of measurement error, I will use simulation studies to demonstrate how effective each modeling approach is at recovering known parameter values from the theory-testing model.

@@

Using the generative causal model in @fig-ideal-dag-theory we can simulate data with a known parameter for the effect, $\beta_1$ of legislator ideology on the outcome $Y$. Then we fit two linear regression models to estimate this parameter. @eq-cont-no-me-model is the naive theory-testing model where $\bar{\theta}_i$ corresponds to a legislator's mean ideal point estimate from the Bayesian IRT measurement model. This is in contrast to the joint measurement theory-testing model in @eq-cont-me-model, which models $\bar{\theta}_i$ as an outcome of the true ideology $\theta_i$ (an unobserved parameter for each observation) and $\sigma^2_{\theta [i]}$ which is the observed variance of the posterior distribution for each groups' ideal point. The parameters $\theta_i$ are also given hyperpriors $\pi$ and $\tau$ for location and scale respectively. The estimates of $\theta_i$ from this simplified measurement model are then used in the linear model which predicts the outcome $y_i$. These, and all other models in this paper, are written in the probabilistic programming language Stan and fit using the No-U-Turn (NUTS) MCMC sampler [@carpenter2017].

@@

$$
\begin{aligned}
  y_i &\sim \text{Normal}(\mu_i, \sigma^2) \\[-10pt]
  \mu_i &= \beta_0 + \beta_1 \bar{\theta}_{i} \\[-10pt]
  \bar{\theta}_{i} &\sim \text{Normal}(\theta_i, \sigma^2_{\theta [i]}) \\[-10pt]
  \theta_i &\sim \text{Normal}(\pi, \tau) \\[-10pt]
  \beta_0, \beta_1 &\sim \text{Normal}(0, 2) \\[-10pt]
  \sigma &\sim \text{Half Student t}(3, 0, 2) \\[-10pt]
  \tau &\sim \text{Half Student t}(3, 0, 2) \\[-10pt]
  \pi &\sim \text{Normal}(0, 1)
\end{aligned}
$$ {#eq-cont-me-model}

@fig-irt-sim-comparison shows how well each model recovers the true parameter value for $\beta_1$: the effect of the true ideal point on the simulated outcome. Each model was fit 40 times across a range of increasing random error levels (shown on the horizontal axis as the correlation between the simulated true ideal point and mean measurement error value approach zero). The mean, and 89% credible interval posterior estimates of each model's $\beta_1$ parameter are plotted with a loess fit. With little-to-no measurement error (left side of the graph), both models reliably recover the true $\beta_1$ value of 1. But as the random measurement error increases, the $\beta_1$ estimates from the naive model from @eq-cont-no-me-model rapidly attenuate towards zero. This is in contrast to the estimates from the joint measurement theory-testing model in @eq-cont-me-model which remain much closer to the true $\beta_1$ value even after there is essentially zero correlation between the true ideal points and means from the ideal points with measurement error.

@@

After the data were generated I fit two models: the ordinary linear regression using only $\bar{\theta}$ values as used previously in the attenuation bias example (@eq-cont-no-me-model), and a modified version of @eq-cont-me-model which substitutes the Normal distribution for the Skew Normal distribution (@eq-cont-skew-model). As before, the key parameter of theoretical interest is $\beta_1$---which, according to the simulated data, should equal zero if unconfounded.

$$
\begin{aligned}
  y_i &\sim \text{Normal}(\mu_i, \sigma^2) \\[-10pt]
  \mu_i &= \beta_0 + \beta_1 \bar{\theta}_{i} \\[-10pt]
  \bar{\theta}_{i} &\sim \text{Skew Normal}(\theta_i, \omega_{\theta[i]}, \alpha_{\theta[i]}) \\[-10pt]
  \theta_i &\sim \text{Normal}(\pi, \tau) \\[-10pt]
  \beta_0, \beta_1 &\sim \text{Normal}(0, 2) \\[-10pt]
  \sigma &\sim \text{Half Student t}(3, 0, 2) \\[-10pt]
  \tau &\sim \text{Half Student t}(3, 0, 2) \\[-10pt]
  \pi &\sim \text{Normal}(0, 1)
\end{aligned}
$$ {#eq-cont-skew-model}

@@

Are ideologically extreme US House incumbents punished electorally? According to the widely cited @canes-wrone2002, the answer is yes. While these authors use interest group scores to measure ideology, they discuss how their results remain the same when using DW-NOMINATE scores of ideology. DW-NOMINATE is a measurement model similar to IRT but without any uncertainty measurements. This makes it unsuitable for replication using the Bayesian method proposed here, so I instead re-estimate all US Representatives' ideology using the popular IRT model from the **pscl** R package [@R-pscl].

@@

$$
\begin{aligned}
  \text{VotePct}_i &\sim \text{Normal}(\mu_i, \sigma^2) \\[-10pt]
  \mu_i &= \beta_0 + \beta_1 \bar{\theta}_{i} + \beta_2\text{R}_i\\[-10pt]
  \bar{\theta}_{i} &\sim \text{Skew Normal}(\theta_i, \omega_{\theta[i]}, \alpha_{\theta[i]}) \\[-10pt]
  \theta_i &\sim \text{Normal}(\pi, \tau) \\[-10pt]
  \beta_0 &\sim \text{Normal}(50, 5) \\[-10pt]
  \beta_1, \beta_2 &\sim \text{Normal}(0, 2) \\[-10pt]
  \sigma &\sim \text{Half Student t}(3, 0, 2) \\[-10pt]
  \tau &\sim \text{Half Student t}(3, 0, 2) \\[-10pt]
  \pi &\sim \text{Normal}(0, 2)
\end{aligned}
$$ {#eq-ree-skew-model}

@@

# Extensions

## Skew-t Distribution

## Measurement Error Validity

@@

Social scientists should therefore be aware of recent computational advancements in Bayesian posterior estimation, particularly the superiority of Hamiltonian Monte Carlo (HMC) samplers over traditional Gibbs samplers (such as the one used in the **pscl** R package). HMC methods enable more accurate and efficient handling of high dimensional parameter spaces [@betancourt2018]---of which IRT models are a prime example. Gibbs samplers also lack the sophisticated suite of diagnostic tools that come with HMC. This means that convergence issues, and therefore poor posterior exploration, may go undetected.

In the same vein as using the best computational sampling methods, this project highlights why latent variables whose measurement models produce uncertainty estimates should be prefered over those that do not. The methodological competitor to IRT models for measuring political ideology is NOMINATE, which uses multidimensional scaling rather than Bayesian estimation. This optimization method does not provide explicit estimates of uncertainty, much less a full posterior distribution of plausible values the latent variable could take. All of NOMINATE's measurement error is therefore in the unobserved category. For the reasons discussed previously, this raises concerns about attenuation bias and/or confounding bias when using NOMINATE values in a theory-testing model.

>>>>

Theory-testing research which uses estimates from measurement models typically reduces the associated posterior distributions down to a single value. In the case of continuous variables, researchers select some statistic of central tendency from each posterior distribution to use in subsequent analyses---such as the mean, median, or mode. This practice necessarily discards information from the full distribution. @fig-measurement-model show four hypothetical posterior distributions that may arise from a Bayesian measurement model. Despite all having the same mean of zero, higher order moments such as variance (top-right), skew (bottom-left), and kurtosis (bottom-right) can create distributions which vary widely.

@@

N <- 100000
tibble(x = c(rnorm(N, sd = 1),
             rnorm(N, sd = 2),
             sn::rsn(N, alpha = -4, omega = 2, xi = 1.5),
             sn::rst(N, xi = -2.75, omega = 2, alpha = 4, nu = 1)),
       y = c(rep("A\nNormal(0, 1)", N), 
             rep("B\nNormal(0, 2)", N), 
             rep("C\nSkew Normal(1.5, 2, -4)", N), 
             rep("D\nSkew t(-2.75, 2, 4, 1)", N))) |> 
  ggplot(aes(x = x)) +
  geom_histogram(color = "white", fill = "darkcyan", alpha = .5, bins = 50) +
  geom_vline(xintercept = 0, linetype = "twodash", color = "black", linewidth = .7) +
  # geom_text(aes(label = y), x = -4, y = 0) +
  facet_wrap(~ y, nrow = 2) +
  xlim(-5, 5) +
  theme_ggdist() +
  labs(y = "", x = "", caption = "Four different measurement model posterior distributions with mean 0") +
  theme(axis.line.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        text = element_text(family = "serif"))
```

@@

How can researchers avoid the issues highlighted above? In short, the measurement process and theory-testing procedure should happen simultaneously in a single model. This is handled straightforwardly using the Bayesian statistical framework, which, unlike the frequentist paradigm, does not draw such a sharp distinction between data and parameters [@mcelreath2020]. We start by specifying the full measurement model, whose posterior distributions for each observation's value of the latent variable are then used as data in the theory-testing model. The stylized version of this joint model is shown in @eq-ex-equation, where $g(\cdot)$ is the measurement model which produces posterior estimates of the latent variable, $\theta_i$ for each observation based on some training data $z$. The posterior estimates for $\theta_i$ from the measurement model $g(\cdot)$ are then used as data in the theory-testing model $f(\theta_i)$ using the outcome of interest $y$.

$$
\begin{aligned}
  y_i &\sim f(\theta_i) \\
  z_i &\sim g(\theta_i)
\end{aligned}
$$ {#eq-ex-equation}

@@

The method developed in this paper overcomes the two problems outlined above by simplifying the measurement model step, $z_i \sim g(\cdot)$ in the joint model. Rather than estimating the latent variable from scratch, I take the posterior distributions already provided from previously fitted measurement models and use those as approximations in the full joint model. The measurement model $g(\cdot)$ becomes a probability distribution function with distributional parameters according to maximum likelihood estimates of the posterior. So if the posterior distribution of the measurement model appears normal, we would use $\bar{\theta}_i \sim N(\theta_i, \sigma^2_{\theta[i]})$. The values $\bar{\theta}_i$ and $\sigma^2_{\theta[i]}$ are estimated from the measurement model's posterior distribution, which allows the true, unobserved, value of the latent variable $\theta_i$ to be estimated for each observation. If the posterior distributions from the measurement model appear skewed, or have thicker tails than a normal distribution, the distributional parameters for these distributions can be used instead. These simplifications faithfully propagate the uncertainty in the outputs of the measurement model to the theory-testing model, while also being computationally tractable and straightforward to implement.

@@

One of the most common measurement models in political science is the Bayesian Item-Response Theory (IRT) model used to measure the ideological leanings of political actors [@clinton2004; @bafumi2005]. These models assume that political ideology is a latent characteristic that lies on a single left-right dimension. Observed actions, such as voting on legislation, are used as training data ($z$ in @eq-ex-equation) to produce a posterior distribution of continuous values for each actor (e.g. member of Congress) on this left-right scale.

Let's say we want to estimate the effect of legislator ideology, $\theta$, on some outcome, $y$. Using the format developed in @eq-ex-equation, @eq-irt shows an example joint measurement and theory-testing model to answer this question using ideology estimates from an IRT model. The measurement model (bottom) predicts whether a legislator voted yes or no on a piece of legislation, $z_j$, using the traditional 2-parameter IRT equation $\Phi(\gamma_j\theta_i + \xi_j)$. Estimates of the parameters $\theta_i$ from this model, are then used as data in the linear regression theory-testing model (top) to estimate $\beta_1$---the coefficient of substantive interest.

$$
\begin{aligned}
  y_i &\sim \text{Normal}(\beta_0 + \beta_1 \theta_i, \sigma^2) \\
  z_{ij} &\sim \text{Bernoulli}[\Phi(\gamma_j\theta_i + \xi_i)] \\
\end{aligned}
$$ {#eq-irt}

@@

$$
\begin{aligned}
  y_i &\sim \text{Normal}(\beta_0 + \beta_1 \theta_i, \sigma^2) \\
  \bar{\theta}_i &\sim \text{Normal}(\theta_i, \sigma^2_{\theta [i]})
\end{aligned}
$$ {#eq-irt-me}

@@

Using the logic of causal graphs, I will discuss two types of measurement model measurement error, and how the joint measurement theory-testing procedure laid out in @eq-ex-equation helps fix them. First I will consider random, or classical, measurement error. In this case the joint model will (in most cases) provide researchers with extra statistical power to test their theory by mitigating issues of attenuation bias. Then we will look at non-random measurement error scenarios, in which the measurement model error introduces confounding in the theory-testing model. I will show how the joint model from @eq-ex-equation ameliorates this confounding bias. For both types of measurement error, I will use simulation studies to demonstrate how effective each modeling approach is at recovering known parameter values from the theory-testing model.

@@

Using the generative causal model in @fig-ideal-dag-theory, we can simulate data with a known parameter for the effect, $\beta_1$ of legislator ideology on the outcome $Y$. Then we fit two linear regression models to estimate this parameter. @eq-cont-no-me-model is the naive theory-testing model where $\bar{\theta}_i$ corresponds to a legislator's mean ideal point estimate from the Bayesian IRT measurement model. This is in contrast to the joint measurement theory-testing model in @eq-cont-me-model, which models $\bar{\theta}_i$ as an outcome of the true ideology $\theta_i$ (an unobserved parameter for each observation) and $\sigma^2_{\theta [i]}$ which is the observed variance of the posterior distribution for each groups' ideal point. The parameters $\theta_i$ are also given hyperpriors $\pi$ and $\tau$ for location and scale respectively. The estimates of $\theta_i$ from this simplified measurement model are then used in the linear model which predicts the outcome $y$. These, and all other models in this paper, are written in the probabilistic programming language Stan and fit using the No-U-Turn (NUTS) MCMC sampler [@carpenter2017].

@@

$$
\begin{aligned}
  y_i &\sim \text{Normal}(\mu_i, \sigma^2) \\[-10pt]
  \mu_i &= \beta_0 + \beta_1 \theta_i \\[-10pt]
  \bar{\theta}_i &\sim \text{Normal}(\theta_i, \sigma^2_{\theta [i]}) \\[-10pt]
  \theta_i &\sim \text{Normal}(\pi, \tau) \\[-10pt]
  \beta_0, \beta_1 &\sim \text{Normal}(0, 2) \\[-10pt]
  \sigma &\sim \text{Half Student t}(3, 0, 2) \\[-10pt]
  \tau &\sim \text{Half Student t}(3, 0, 2) \\[-10pt]
  \pi &\sim \text{Normal}(0, 1)
\end{aligned}
$$ {#eq-cont-me-model}

@fig-irt-sim-comparison shows how well each model recovers the true parameter value for $\beta_1$: the effect of the true ideal point on the simulated outcome. Each model was fit 40 times across a range of increasing values for $\sigma^2_{\theta [i]}$, thereby increasing random error in the independent variable (shown on the horizontal axis as the correlation between the simulated true ideal point and mean measurement error value approach zero). The mean, and 89% credible interval posterior estimates of each model's $\beta_1$ parameter are plotted with a loess fit. With little-to-no measurement error (left side of the graph), both models reliably recover the true $\beta_1$ value of 1. But as the random measurement error increases, the $\beta_1$ estimates from the naive model from @eq-cont-no-me-model rapidly attenuate towards zero. This is in contrast to the estimates from the joint measurement theory-testing model in @eq-cont-me-model which remain much closer to the true $\beta_1$ value even after there is essentially zero correlation between the true ideal points and means from the ideal points with measurement error.

@@

After the data are generated, we fit two models: the ordinary linear regression using only $\bar{\theta}$ values as used previously in the attenuation bias example (@eq-cont-no-me-model), and a modified version of @eq-cont-me-model which substitutes the Normal distribution for the Skew Normal distribution (@eq-cont-skew-model). As before, the key parameter of theoretical interest is $\beta_1$---which, according to the simulated data, should equal zero if unconfounded.

$$
\begin{aligned}
  y_i &\sim \text{Normal}(\mu_i, \sigma^2) \\[-10pt]
  \mu_i &= \beta_0 + \beta_1 \theta_i \\[-10pt]
  \bar{\theta}_{i} &\sim \text{Skew Normal}(\theta_i, \omega_{\theta[i]}, \alpha_{\theta[i]}) \\[-10pt]
  \theta_i &\sim \text{Normal}(\pi, \tau) \\[-10pt]
  \beta_0, \beta_1 &\sim \text{Normal}(0, 2) \\[-10pt]
  \sigma &\sim \text{Half Student t}(3, 0, 2) \\[-10pt]
  \tau &\sim \text{Half Student t}(3, 0, 2) \\[-10pt]
  \pi &\sim \text{Normal}(0, 1)
\end{aligned}
$$ {#eq-cont-skew-model}

@@

Are ideologically extreme US House incumbents punished electorally? According to the widely cited @canes-wrone2002, the answer is yes. While these authors use interest group scores to measure ideology, they discuss how their results remain the same when using DW-NOMINATE scores of ideology. DW-NOMINATE is a measurement model similar to IRT, but whose uncertainty measurements are only given in the form of bootstrapped standard errors [@carroll2009]. These standard error estimates could potentially be used as $\sigma^2_{\theta[i]}$ in the classical measurement error model $\bar{\theta}_i \sim N(\theta_i, \sigma^2_{\theta[i]})$ since Bayesian posterior standard deviations and frequentist standard errors share similar qualities. But standard errors do not provide any information about error skewness. This makes them unsuitable for replication using the Bayesian method proposed here, so I instead re-estimate all US Representatives' ideology using the popular IRT model from the **pscl** R package [@R-pscl].

@@

$$
\begin{aligned}
  \text{VotePct}_i &\sim \text{Normal}(\mu_i, \sigma^2) \\[-10pt]
  \mu_i &= \beta_0 + \beta_1 \theta_i + \beta_2\text{R}_i\\[-10pt]
  \bar{\theta}_{i} &\sim \text{Skew Normal}(\theta_i, \omega_{\theta[i]}, \alpha_{\theta[i]}) \\[-10pt]
  \theta_i &\sim \text{Normal}(\pi, \tau) \\[-10pt]
  \beta_0 &\sim \text{Normal}(50, 5) \\[-10pt]
  \beta_1, \beta_2 &\sim \text{Normal}(0, 2) \\[-10pt]
  \sigma &\sim \text{Half Student t}(3, 0, 2) \\[-10pt]
  \tau &\sim \text{Half Student t}(3, 0, 2) \\[-10pt]
  \pi &\sim \text{Normal}(0, 2)
\end{aligned}
$$ {#eq-ree-skew-model}

@@

# Measurement Error Validity

@@

Social scientists should therefore be aware of recent computational advancements in Bayesian posterior estimation, particularly the superiority of Hamiltonian Monte Carlo (HMC) samplers over traditional Gibbs samplers (such as the one used in the **pscl** R package). HMC methods enable more accurate and efficient handling of high dimensional parameter spaces [@betancourt2018]---of which IRT models are a prime example. Gibbs samplers also lack the sophisticated suite of diagnostic tools that come with HMC. This means that convergence issues, and therefore poor posterior exploration, may go undetected. These makes the observed error estimates in the posterior less trustworthy.

In the same vein as using the best computational sampling methods, this project highlights why latent variables whose measurement models produce rich uncertainty estimates should be preferred over those that do not. The methodological competitor to IRT models for measuring political ideology is DW-NOMINATE, which uses multidimensional scaling rather than Bayesian estimation. This optimization method does not provide explicit estimates of uncertainty, much less a full posterior distribution of plausible values the latent variable could take. All of DW-NOMINATE's uncertainty estimates must come from bootstrap procedures which produce, at best, only standard errors. This means that more of this model's measurement error is in the unobserved category, which in turn raises concerns about attenuation bias and/or confounding bias when using DW-NOMINATE values in a theory-testing model.

# Conclusion

This project highlights the problems associated with ignoring measurement error when testing theories which rely on measurement model variables. Doing so will often lead to attenuation bias, which could lead to the mistaken conclusion that no relationship between the independent variable and dependent variable exist, when in fact it does. And in even more problematic cases, ignoring measurement model error can introduce confounding bias into the theory-testing analysis. The method proposed here: to simultaneously estimate the theory-testing model and measurement model at once, helps fix these two issues to the extent that the posterior distribution of the measurement model is valid.
